---
title: "DS311 - Basic R Lab Exercise"
subtitle: "R Lab Exercise"
author: "Your Name"
date: "1/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

What is your name?

```{r}
# Write your code here!s
print()
```

***

# Question 2

How many classes you are taking this semester?

```{r}
# Write your code here!
print()
```

***

# Question 3

Calculate the following problems:

a. 2 + 4 x 6

```{r}
# Write your code here!


print()
```

b. $\sqrt{9}$

```{r}
# Write your code here!


print()
```

c. $^6/_3$

```{r}
# Write your code here!


print()
```
***

## Question 4

Calculate the following problem step-by-step.

[5 - (6 x 3) + (4 / 2) x 10] + (20 / 5) - 10

```{r}
# Write your code here!


print()
```

***

## Question 5
Write the code that accomplish the following tasks:

Part a: Assign 4 to variable x

Part b: Assign 12 to variable y

Part c: Print both x and y to check their values

Part d: Divide y by x and assign it to variable z

part e: Print a statement to report your answer in Part d.

```{r}
# Write your code here!
# Part a


# Part b


# Part c


# Part d


# Part e
print(paste("y divided by x is equal to ", "Ans"))


# Do not need to change the following code!
if (exists("x") == TRUE | exists("y") == TRUE | exists("z") == TRUE){
  if (x == 4 & y == 12 & z == 3) {
  print("Congratulation!!  You completed the first activity in this class!!")
  } else {
    print("Sorry, you got it wrong!")
  }
} else {
  print("You did not complete the last problem!")
}
```

